ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel

### 2025-01-17 paper/Manuscript-Data-Package.tex
inclusion of section for python integration and how to use this package

\begin{abstract}
This paper describes \{ActiveCA\}, and open data product with Canadian
time use data. \{ActiveCA\} is an \texttt{R} data package that contains
analysis-ready data related to active travel spanning almost 40 years,
extracted from cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005), 24
(2010), and 29 (2015) of Canada's General Social Survey. Active travel
is characterized by mode, with walking being part of every cycle and
bicycling starting in 1992. The attributes of active trips are the types
of locations of origins and destinations, the duration of trips, and
episode weights for expanding the trips to population-wide estimates.
Based on the year of the survey, a variety of locations are coded. In
earlier cycles, these include home, work or school, and other's home,
whereas in later cycles these are augmented with locations such as
grocery stores, restaurants, outdoor destinations, and others. The
geographical resolution includes the province and whether the episode
was in an urban or rural setting.
\end{abstract}

@@

\section{Introduction}\label{introduction}

@@

\section{General Social Survey (GSS)
collection}\label{general-social-survey-gss-collection}

@@

\subsection{The Main File}\label{the-main-file}

The main file of the time use survey compiles a large array of
aggregated data, summarizing the answers to the questionnaire that
describe households and individuals, as well as derived variables that
summarize the respondents' use of time use across different activities,
locations, and social interactions. This file documents the time and
duration that respondents allocate to each activity and location. The
Main File provides a overview of daily routines and social dynamics, not
focusing on individual activity episodes. Additionally, this file
categorizes activities into bigger groups and subcategories,
facilitating the data's analytical utility with additional metrics such
as total transit time, time spent with household members, and counts of
activities and episodes.

Table \ref{tab:main-2015-unprocessed} shows the first ten rows and first
six variables of the GSS PUMF 2015 main file (Cycle 29). Each row in the
table correspond to a survey respondent, while the columns refer the
following information: record identification (\texttt{PUMFID}), the
person's weight (\texttt{WGHT\_PER}), the month the survey data was
collected (\texttt{SURVMNTH}), the respondent's age group
(\texttt{AGEGR10}), the respondent's sex (\texttt{SEX}), and the
respondent's marital status (\texttt{MARSTAT}).

The main file of the 2015 GSS surveys includes a total of 17,390
respondents, representing 29,766,399 individuals and 848 variables. For
discrete variables, Statistics Canada has assigned specific codes to the
possible values, with each code accompanied by a label. For instance, in
the case of the variable \texttt{SURVMNTH}, a value of \texttt{1} means
\texttt{January\ 2016}, 2 means \texttt{February\ 2016}, 3 corresponds
to \texttt{March\ 2016}, and so on.

@@

\begin{ThreePartTable}
\begin{TableNotes}
\item \textit{Note: } 
\item Legend: `PUMFID`: record identification. `WGHT\_PER`:  person weight. `SURVMNTH`: survey month of data collection. `AGEGR10`: age group of the respondent. `SEX`: sex of the respondent. `MARSTAT`: marital status of the respondent.
\end{TableNotes}
\begin{longtable}[t]{cccccc}
\caption{\label{tab:gss-main-file-2015}\label{tab:main-2015-unprocessed}Visualization of the first ten lines and first six columns of the main file of the 2015 GSS.}\\
\toprule
PUMFID & WGHT\_PER & SURVMNTH & AGEGR10 & SEX & MARSTAT\\
\midrule
10000 & 616.6740 & 7 & 5 & 1 & 5\\
10001 & 8516.6140 & 7 & 5 & 1 & 1\\
10002 & 371.7520 & 1 & 4 & 2 & 1\\
10003 & 1019.3135 & 3 & 6 & 2 & 5\\
10004 & 1916.0708 & 9 & 2 & 1 & 6\\
\addlinespace
10005 & 1952.2015 & 4 & 1 & 1 & 6\\
10006 & 5761.5528 & 8 & 1 & 1 & 6\\
10007 & 466.0426 & 6 & 5 & 2 & 3\\
10008 & 2479.2991 & 2 & 2 & 2 & 1\\
10009 & 1436.1641 & 8 & 6 & 1 & 3\\
\bottomrule
\insertTableNotes
\end{longtable}
\end{ThreePartTable}
\endgroup{}

\subsection{The Episode File}\label{the-episode-file}

The episode is a much bigger file that records detailed data for each
activity episode reported by respondents. Each episode represents a
single activity and its duration, and the sum of all episodes throughout
the day adds up to 24 hours. Each entry in this file includes the start
and end times of the activity, the duration, location, and accompanying
social context, informing when and where activities occurred and with
whom. The focus of the Episode File is not on the characteristics of the
respondents, but on the characteristics of the activities, and the data
are structured around the numerous activity instances that compose a day
of the respondent. Although respondent-specific characteristics are not
included within the episode file, it is possible to link the main file
and the episode file by using a key present in both the Main and Episode
Files.

Similar to Table \ref{tab:main-2015-unprocessed}, which displayed an
example of the main file structure, Table \ref{tab:ep-2015-unprocessed}
presents the first seven episodes for the record identification number
\texttt{10041} and some variables from the GSS PUMF 2015 episode file
(Cycle 29). Each row in the table corresponds to an episode associated
with the specified record identification (\texttt{PUMFID\ =\ 10041}),
including the episode's weight (\texttt{WGHT\_EPI}), episode number
(\texttt{EPINO}), activity code (\texttt{TUI\_01}), episode duration
(\texttt{DURATION}), and episode location (\texttt{LOCATION}).

In total, the episode file of the 2015 GSS surveys contains 274,108
records, representing 461,837,622 episodes and 527 variables detailing
the episodes. Similar to the main file, Statistics Canada has created
codes for the discrete variables, with each value corresponding to a
label.

@@

\section{Data extraction}\label{data-extraction}

For each selected cycle of the GSS surveys, we reviewed the episode
files to identify episodes of movement that involved walking or cycling.
This allowed us to select the activities immediately before and after
the movement episode. After that, we labeled the code variables with
their appropriate descriptions, identifying each origin and destination,
mode of travel, time spent in the active trip, as well the province and
urban classification of the episode.

@@

For each GSS main file, we selected socioeconomic variables to help
profile the individuals engaged in active trip episodes. We included key
socioeconomic variables such as the respondent's age group, sex, marital
status, and number of children, among others. After this, as we did with
the episode files, we labeled the code variables with their appropriate
descriptions.

\section{\texorpdfstring{How to use
\texttt{ActiveCA}}{How to use ActiveCA}}\label{how-to-use-activeca}

This section presents some potential applications of the
\texttt{ActiveCA} R package. In fact, we expect that the application of
this package to extend beyond our pre-imagined range of uses.

To start the demonstration, it is first necessary to install the
package. The code chunk below demonstrates how to install the
\texttt{ActiveCA} package in R:

@@

After installing the \texttt{ActiveCA} R package, you can load it using
the following code:

@@

For instance, if someone wants to calculate the percentage of
respondents from the 2015 GSS survey with active travel episodes, it is
necessary to account for the person's weights. In 2015, the weight
variable is represented by \texttt{WGHT\_PER.} The code chunk below
demonstrates how to obtain the percentage of people with active travel
episodes by age group. It uses the dplyr package to manipulate the data.

The process begins by creating a dataset that sums the population by age
group. Then, it joins the 2015 episodes with the 2015 main file. Note
that in both operations, the code sums the person weight variable
(\texttt{WGHT\_PER}) to obtain the correct population values. After
this, a new dataset, \texttt{Active\_percentage}, is created by merging
both previous datasets. The percentage of active travel episodes by age
group is calculated by dividing the total population by the population
with active trip episodes, then multiplying by 100.

@@

\subsection{Processed data set}\label{processed-data-set}

Table \ref{tab:processed-obs} displays the total number of observations
processed for main and episode files. For the main files, a total of
89,331 observations were processed, referring to all observations from
the Time Use Surveys from 1986 to 2015, that together represents more of
149,389,839 respondents (29,766,399 for 2015, 28,075,610 for 2010,
26,095,819 for 2005, 24,260,137 for 1998, 21,294,313 for 1992, and for
2005, 24,260,137 for 1998, and 21,294,313 for 1986). Table X also
presents the total cases of active trips episodes identified. In total
21,748 observations with register of active travel activity. Together,
these observations account for NA episodes (0 for 2015, 0 for 2010, 0
for 2005, 0 for 1998, 0 for 1992, and for 2005, 0 for 1998, and 0 for
1986).

@@

\begin{longtable}[t]{lrrr}
\caption{\label{tab:table_df_processed}\label{tab:processed-obs}Total number and weighted sum of observations processed.}\\
\toprule
\multicolumn{2}{c}{ } & \multicolumn{2}{c}{Observations} \\
\cmidrule(l{3pt}r{3pt}){3-4}
Survey & Year & Count & Weighted\\
\midrule
 & 2015 & 17390 & 29766399\\
\nopagebreak
 & 2010 & 15390 & 28075610\\
\nopagebreak
 & 2005 & 19597 & 26095819\\
\nopagebreak
 & 1998 & 10749 & 24260137\\
\nopagebreak
 & 1992 & 9815 & 21294313\\
\nopagebreak
\multirow[t]{-6}{*}{\raggedright\arraybackslash Main} & 1986 & 16390 & 19897562\\
\cmidrule{1-4}\pagebreak[0]
 & 2015 & 3496 & 6634387\\
\nopagebreak
 & 2010 & 4615 & 8516753\\
\nopagebreak
 & 2005 & 5866 & 7583838\\
\nopagebreak
 & 1998 & 1789 & 3606987\\
\nopagebreak
 & 1992 & 1635 & \\
\nopagebreak
\multirow[t]{-6}{*}{\raggedright\arraybackslash Episode} & 1986 & 4347 & 8241196\\
\bottomrule
\end{longtable}
\endgroup{}

Table \ref{tab:main-2015-processed} shows the first ten rows and first
six variables of the GSS PUMF 2015 main file (Cycle 29), displayed in
\ref{tab:main-2015-unprocessed} before our processing. Table
\ref{tab:ep-2015-processed} presents the walking episodes for the record
identification number \texttt{10041} from the GSS PUMF 2015 episode file
(Cycle 29), previously displayed in Table \ref{tab:ep-2015-unprocessed}.
Only active travel episodes appear in Table \ref{tab:ep-2015-processed}
since the observations were filtered to select cases with walking or
cycling episodes. For both cases, Tables \ref{tab:main-2015-processed}
and \ref{tab:ep-2015-processed} contain labeled variables, facilitating
the interpretation of the data.

@@

\begin{ThreePartTable}
\begin{TableNotes}
\item \textit{Note: } 
\item Legend: `PUMFID`: record identification. `WGHT\_PER`:  person weight. `SURVMNTH`: survey month of data collection. `AGEGR10`: age group of the respondent. `SEX`: sex of the respondent. `MARSTAT`: marital status of the respondent.
\end{TableNotes}
\begin{longtable}[t]{cccccc}
\caption{\label{tab:gss-processed-file-2015}\label{tab:main-2015-processed}Visualization of the first ten lines and first six columns of the main file of the 2015 GSS.}\\
\toprule
PUMFID & WGHT\_PER & SURVMNTH & AGEGR10 & SEX & MARSTAT\\
\midrule
10000 & 616.6740 & July & 55 to 64 years & Male & Divorced\\
10001 & 8516.6140 & July & 55 to 64 years & Male & Married\\
10002 & 371.7520 & January & 45 to 54 years & Female & Married\\
10003 & 1019.3135 & March & 65 to 74 years & Female & Divorced\\
10004 & 1916.0708 & September & 25 to 34 years & Male & Single, never married\\
\addlinespace
10005 & 1952.2015 & April & 15 to 24 years & Male & Single, never married\\
10006 & 5761.5528 & August & 15 to 24 years & Male & Single, never married\\
10007 & 466.0426 & June & 55 to 64 years & Female & Widowed\\
10008 & 2479.2991 & February & 25 to 34 years & Female & Married\\
10009 & 1436.1641 & August & 65 to 74 years & Male & Widowed\\
\bottomrule
\insertTableNotes
\end{longtable}
\end{ThreePartTable}
\endgroup{}

@@

\begin{ThreePartTable}
\begin{TableNotes}
\item \textit{Note: } 
\item Legend: `PUMFID`: record identification. `EPINO`: episode number. WGHT\_EPI: episode's weight. TUI\_01: activity code. DURATION: episode's duration. LOCATION: episode's location.
\end{TableNotes}
\begin{longtable}[t]{rrlrlll}
\caption{\label{tab:gss-processed-file-2015}\label{tab:ep-2015-processed}Visualization of active travel episodes of the record number `10041` for the 2015 GSS survey.}\\
\toprule
PUMFID & WGHT\_EPI & Activity & Duration & Origin & Destination & Mode\\
\midrule
10041 & 1353.818 & Transport to or from activity & 15 & Home & Home & Walking\\
\bottomrule
\insertTableNotes
\end{longtable}
\end{ThreePartTable}
\endgroup{}

\subsection{Descriptive statistics}\label{descriptive-statistics}

Considering GSS Cycles analyzed, we identified 21,748 episodes that
recorded active travel episodes, with trip duration ranging from 0 to
900 minutes, to twelve different destinations. \texttt{ActiveCA}
includes all these episodes ready for analysis. Table \ref{tab:table-01}
presents descriptive statistics on walking and cycling trips between
1986 and 2015, with measures of trip duration in minutes: maximum (max),
mean, median, and minimum (min). The 1986 survey did not include bicycle
trips.

@@

\begin{longtable}[t]{>{}llcccccc}
\caption{\label{tab:table-01}\label{tab:table-01}Descriptive statistics for episodes with active transport records}\\
\toprule
\multicolumn{2}{c}{ } & \multicolumn{6}{c}{Year} \\
\cmidrule(l{3pt}r{3pt}){3-8}
Mode & Statistic & 1986 & 1992 & 1998 & 2005 & 2010 & 2015\\
\midrule
 & Maximum & 660 & 300 & 255 & 515 & 480 & 900\\
\nopagebreak
 & Mean & 21 &  & 12 & 12 & 13 & 18\\
\nopagebreak
 & Median & 15 & 10 & 5 & 10 & 10 & 10\\
\nopagebreak
 & Minimum & 1 & 1 & 1 & 0 & 0 & 5\\
\nopagebreak
\multirow[t]{-5}{*}{\raggedright\arraybackslash \textbf{Walking}} & Standard deviation & 31 & 25 & 17 & 16 & 17 & 27\\
\cmidrule{1-8}\pagebreak[0]
 & Maximum &  & 240 & 90 & 180 & 153 & 120\\
\nopagebreak
 & Mean &  & 28 & 24 & 20 & 19 & 25\\
\nopagebreak
 & Median &  & 15 & 15 & 15 & 10 & 20\\
\nopagebreak
 & Minimum &  & 5 & 2 & 1 & 1 & 5\\
\nopagebreak
\multirow[t]{-5}{*}{\raggedright\arraybackslash \textbf{Cycling}} & Standard deviation &  & 36 & 18 & 18 & 23 & 20\\
\bottomrule
\end{longtable}
\endgroup{}

@@

\begin{ThreePartTable}
\begin{TableNotes}
\item \textit{Note: } 
\item 'Min' denotes the minimum time to reach the destination; 'Max' denotes the maximum time to reach the destination; '(\%)' indicates a percentage of the total time to reach the destination; 'Med' refers to the median time to reach the destination
\end{TableNotes}
\begin{longtable}[t]{ccccc>{}c|ccc>{}c|cccc}
\caption{\label{tab:table-02}\label{tab:table-02}Comparison of travel statistics by mode and destination: 1986, 1992, 1998}\\
\toprule
\multicolumn{2}{c}{ } & \multicolumn{4}{c}{1986} & \multicolumn{4}{c}{1992} & \multicolumn{4}{c}{1998} \\
\cmidrule(l{3pt}r{3pt}){3-6} \cmidrule(l{3pt}r{3pt}){7-10} \cmidrule(l{3pt}r{3pt}){11-14}
\multicolumn{1}{c}{\textbf{Destination}} & \multicolumn{1}{c}{\textbf{Mode}} & \multicolumn{1}{c}{\textbf{Min}} & \multicolumn{1}{c}{\textbf{Med}} & \multicolumn{1}{c}{\textbf{Max}} & \multicolumn{1}{c}{\textbf{(\%)}} & \multicolumn{1}{c}{\textbf{Min}} & \multicolumn{1}{c}{\textbf{Med}} & \multicolumn{1}{c}{\textbf{Max}} & \multicolumn{1}{c}{\textbf{(\%)}} & \multicolumn{1}{c}{\textbf{Min}} & \multicolumn{1}{c}{\textbf{Med}} & \multicolumn{1}{c}{\textbf{Max}} & \multicolumn{1}{c}{\textbf{(\%)}}\\
\midrule
 & Home &  &  &  &  & 5 & 20 & 240 & 55.7 & 2 & 15 & 90 & 51.6\\
\nopagebreak
 & Other's home &  &  &  &  & 5 & 10 & 145 & 19.7 & 2 & 10 & 80 & 15.7\\
\nopagebreak
\multirow[t]{-3}{*}{\centering\arraybackslash Cycling} & Work or school &  &  &  &  & 5 & 15 & 45 & 24.7 & 5 & 25 & 75 & 32.8\\
\cmidrule{1-14}\pagebreak[0]
 & Home & 1 & 15 & 330 & 46.3 & 1 & 15 & 300 &  & 1 & 5 & 255 & 51.9\\
\nopagebreak
 & Other's home & 1 & 10 & 660 & 42.6 & 1 & 5 & 135 &  & 1 & 5 & 120 & 27.8\\
\nopagebreak
\multirow[t]{-3}{*}{\centering\arraybackslash Walking} & Work or school & 1 & 10 & 450 & 11.1 & 2 & 10 & 60 &  & 1 & 10 & 75 & 20.3\\
\bottomrule
\insertTableNotes
\end{longtable}
\end{ThreePartTable}
\endgroup{}

@@

\texttt{ActiveCA} also enables obtaining insights from the main
processed files. Figure \ref{fig:figure-stress} present how the level of
stress varied among respondents depending on their marital status in
2015. According to this plot, married respondents reported the highest
level of stress, with 17\% of possible cases.

>>>>

\begin{abstract}
This paper describes \{ActiveCA\}, an open data product with Canadian
time use data. \{ActiveCA\} is an \texttt{R} data package that contains
analysis-ready data related to active travel spanning almost 40 years,
extracted from cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005), 24
(2010), and 29 (2015) of Canada's General Social Survey. Active travel
is characterized by mode, with walking being part of every cycle and
bicycling starting in 1992. The attributes of active trips are the types
of locations of origins and destinations, the duration of trips, and
episode weights for expanding the trips to population-wide estimates.
Based on the year of the survey, a variety of locations are coded. In
earlier cycles, these include home, work or school, and other's home,
whereas in later cycles these are augmented with locations such as
grocery stores, restaurants, outdoor destinations, and others. The
geographical resolution includes the province and whether the episode
was in an urban or rural setting.
\end{abstract}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

\hypertarget{general-social-survey-gss-collection}{%
\section{General Social Survey (GSS)
collection}\label{general-social-survey-gss-collection}}

@@

\hypertarget{the-main-file}{%
\subsection{The Main File}\label{the-main-file}}

The Main File of the time use survey compiles a large array of
aggregated data, summarizing the answers to the questionnaire that
describe households and individuals, as well as derived variables that
summarize the respondents' use of time use across different activities,
locations, and social interactions. This file documents the time and
duration that respondents allocate to each activity and location. The
Main File provides a overview of daily routines and social dynamics, not
focusing on individual activity episodes. Additionally, this file
categorizes activities into bigger groups and subcategories,
facilitating the data's analytical utility with additional metrics such
as total transit time, time spent with household members, and counts of
activities and episodes.

Table \ref{tab:main-2015-unprocessed} shows the first ten rows and first
six variables of the GSS PUMF 2015 Main File (Cycle 29). Each row in the
table correspond to a survey respondent, while the columns refer the
following information: record identification (\texttt{PUMFID}), the
person's weight (\texttt{WGHT\_PER}), the month the survey data was
collected (\texttt{SURVMNTH}), the respondent's age group
(\texttt{AGEGR10}), the respondent's sex (\texttt{SEX}), and the
respondent's marital status (\texttt{MARSTAT}).

The Main File of the 2015 GSS surveys includes a total of 17,390
respondents, representing 29,766,399 individuals and 848 variables. For
discrete variables, Statistics Canada has assigned specific codes to the
possible values, with each code accompanied by a label. For instance, in
the case of the variable \texttt{SURVMNTH}, a value of \texttt{1} means
\texttt{January\ 2016}, 2 means \texttt{February\ 2016}, 3 corresponds
to \texttt{March\ 2016}, and so on.

@@

\begin{ThreePartTable}
\begin{TableNotes}
\item \textit{Note: } 
\item Legend: `PUMFID`: record identification. `WGHT\_PER`:  person weight. `SURVMNTH`: survey month of data collection. `AGEGR10`: age group of the respondent. `SEX`: sex of the respondent. `MARSTAT`: marital status of the respondent.
\end{TableNotes}
\begin{longtable}[t]{cccccc}
\caption{\label{tab:gss-main-file-2015}\label{tab:main-2015-unprocessed}Visualization of the first ten lines and first six columns of the Main File of the 2015 GSS.}\\
\toprule
PUMFID & WGHT\_PER & SURVMNTH & AGEGR10 & SEX & MARSTAT\\
\midrule
10000 & 616.6740 & 7 & 5 & 1 & 5\\
10001 & 8516.6140 & 7 & 5 & 1 & 1\\
10002 & 371.7520 & 1 & 4 & 2 & 1\\
10003 & 1019.3135 & 3 & 6 & 2 & 5\\
10004 & 1916.0708 & 9 & 2 & 1 & 6\\
\addlinespace
10005 & 1952.2015 & 4 & 1 & 1 & 6\\
10006 & 5761.5528 & 8 & 1 & 1 & 6\\
10007 & 466.0426 & 6 & 5 & 2 & 3\\
10008 & 2479.2991 & 2 & 2 & 2 & 1\\
10009 & 1436.1641 & 8 & 6 & 1 & 3\\
\bottomrule
\insertTableNotes
\end{longtable}
\end{ThreePartTable}
\endgroup{}

\hypertarget{the-episode-file}{%
\subsection{The Episode File}\label{the-episode-file}}

The episode is a much bigger file that records detailed data for each
activity episode reported by respondents. Each episode represents a
single activity and its duration, and the sum of all episodes throughout
the day adds up to 24 hours. Each entry in this file includes the start
and end times of the activity, the duration, location, and accompanying
social context, informing when and where activities occurred and with
whom. The focus of the Episode File is not on the characteristics of the
respondents, but on the characteristics of the activities, and the data
are structured around the numerous activity instances that compose a day
of the respondent. Although respondent-specific characteristics are not
included within the Episode File, it is possible to link the Main File
and the Episode File by using a key present in both the Main and Episode
Files.

Similar to Table \ref{tab:main-2015-unprocessed}, which displayed an
example of the Main File structure, Table \ref{tab:ep-2015-unprocessed}
presents the first seven episodes for the record identification number
\texttt{10041} and some variables from the GSS PUMF 2015 Episode File
(Cycle 29). Each row in the table corresponds to an episode associated
with the specified record identification (\texttt{PUMFID\ =\ 10041}),
including the episode's weight (\texttt{WGHT\_EPI}), episode number
(\texttt{EPINO}), activity code (\texttt{TUI\_01}), episode duration
(\texttt{DURATION}), and episode location (\texttt{LOCATION}).

In total, the Episode File of the 2015 GSS surveys contains 274,108
records, representing 461,837,622 episodes and 527 variables detailing
the episodes. Similar to the Main File, Statistics Canada has created
codes for the discrete variables, with each value corresponding to a
label.

@@

\hypertarget{data-extraction}{%
\section{Data extraction}\label{data-extraction}}

For each selected cycle of the GSS surveys, we reviewed the Episode
Files to identify episodes of movement that involved walking or cycling.
This allowed us to select the activities immediately before and after
the movement episode. After that, we labeled the code variables with
their appropriate descriptions, identifying each origin and destination,
mode of travel, time spent in the active trip, as well the province and
urban classification of the episode.

@@

For each GSS Main File, we selected socioeconomic variables to help
profile the individuals engaged in active trip episodes. We included key
socioeconomic variables such as the respondent's age group, sex, marital
status, and number of children, among others. After this, as we did with
the Episode Files, we labeled the code variables with their appropriate
descriptions.

\hypertarget{how-to-use-activeca}{%
\section{How to use \{ActiveCA\}}\label{how-to-use-activeca}}

This section presents some potential applications of the \{ActiveCA\} R
package. In fact, we expect that the application of this package to
extend beyond our pre-imagined range of uses.

To start the demonstration, it is first necessary to install the
package. The code chunk below demonstrates how to install the
\{ActiveCA\} package in R:

@@

After installing the \{ActiveCA\} R package, you can load it using the
following code:

@@

For instance, if someone wants to calculate the percentage of
respondents from the 2015 GSS survey with active travel episodes, it is
necessary to account for the person's weights. In 2015, the weight
variable is represented by \texttt{WGHT\_PER.} The code chunk below
demonstrates how to obtain the percentage of people with active travel
episodes by age group. It uses the \texttt{dplyr}package
\citep{dplyr2023} to manipulate the data.

The process begins by creating a dataset that sums the population by age
group. Then, it joins the 2015 episodes with the 2015 Main File. Note
that in both operations, the code sums the person weight variable
(\texttt{WGHT\_PER}) to obtain the correct population values. After
this, a new dataset, \texttt{Active\_percentage}, is created by merging
both previous datasets. The percentage of active travel episodes by age
group is calculated by dividing the total population by the population
with active trip episodes, then multiplying by 100 and rounding with 2
decimal places.

@@

\hypertarget{processed-data-set}{%
\subsection{Processed data set}\label{processed-data-set}}

Table \ref{tab:processed-obs} displays the total number of records
processed for Main and Episode Files. For the Main Files, a total of
89,331 records were processed, referring to all records from the Time
Use Surveys from 1986 to 2015, that together represents more of
149,389,839 respondents (29,766,399 for 2015, 28,075,610 for 2010,
26,095,819 for 2005, 24,260,137 for 1998, 21,294,313 for 1992, and for
2005, 24,260,137 for 1998, and 21,294,313 for 1986). It also presents
the total cases of active trips episodes identified. In total 21,748
records with register of active travel activity. Together, these records
account for 38,275,077 episodes.

@@

\begin{longtable}[t]{lrrr}
\caption{\label{tab:table_df_processed}\label{tab:processed-obs}Total number and weighted sum of records processed.}\\
\toprule
\multicolumn{2}{c}{ } & \multicolumn{2}{c}{Observations} \\
\cmidrule(l{3pt}r{3pt}){3-4}
Survey & Year & Count & Weighted\\
\midrule
 & 2015 & 17390 & 29766399\\
\nopagebreak
 & 2010 & 15390 & 28075610\\
\nopagebreak
 & 2005 & 19597 & 26095819\\
\nopagebreak
 & 1998 & 10749 & 24260137\\
\nopagebreak
 & 1992 & 9815 & 21294313\\
\nopagebreak
\multirow[t]{-6}{*}{\raggedright\arraybackslash Main} & 1986 & 16390 & 19897562\\
\cmidrule{1-4}\pagebreak[0]
 & 2015 & 3496 & 6634387\\
\nopagebreak
 & 2010 & 4615 & 8516753\\
\nopagebreak
 & 2005 & 5866 & 7583838\\
\nopagebreak
 & 1998 & 1789 & 3606987\\
\nopagebreak
 & 1992 & 1635 & 3691918\\
\nopagebreak
\multirow[t]{-6}{*}{\raggedright\arraybackslash Episode} & 1986 & 4347 & 8241196\\
\bottomrule
\end{longtable}
\endgroup{}

Table \ref{tab:main-2015-processed} shows the first ten rows and first
six variables of the GSS PUMF 2015 Main File (Cycle 29), displayed in
\ref{tab:main-2015-unprocessed} before our processing. Table
\ref{tab:ep-2015-processed} presents the walking episodes for the record
identification number \texttt{10041} from the GSS PUMF 2015 Episode File
(Cycle 29), previously displayed in Table \ref{tab:ep-2015-unprocessed}.
Only the unique active travel episode appear in Table
\ref{tab:ep-2015-processed} since the records were filtered to select
cases with walking or cycling episodes. For both cases, Tables
\ref{tab:main-2015-processed} and \ref{tab:ep-2015-processed} contain
labeled variables, facilitating the interpretation of the data.

@@

\begin{ThreePartTable}
\begin{TableNotes}
\item \textit{Note: } 
\item Legend: `PUMFID`: record identification. `WGHT\_PER`:  person weight. `SURVMNTH`: survey month of data collection. `AGEGR10`: age group of the respondent. `SEX`: sex of the respondent. `MARSTAT`: marital status of the respondent.
\end{TableNotes}
\begin{longtable}[t]{cccccc}
\caption{\label{tab:gss-processed-file-2015}\label{tab:main-2015-processed}Visualization of the first ten lines and first six columns of the Main File of the 2015 GSS.}\\
\toprule
PUMFID & WGHT\_PER & SURVMNTH & AGEGR10 & SEX & MARSTAT\\
\midrule
10000 & 616.6740 & July & 55 to 64 years & Male & Divorced\\
10001 & 8516.6140 & July & 55 to 64 years & Male & Married\\
10002 & 371.7520 & January & 45 to 54 years & Female & Married\\
10003 & 1019.3135 & March & 65 to 74 years & Female & Divorced\\
10004 & 1916.0708 & September & 25 to 34 years & Male & Single, never married\\
\addlinespace
10005 & 1952.2015 & April & 15 to 24 years & Male & Single, never married\\
10006 & 5761.5528 & August & 15 to 24 years & Male & Single, never married\\
10007 & 466.0426 & June & 55 to 64 years & Female & Widowed\\
10008 & 2479.2991 & February & 25 to 34 years & Female & Married\\
10009 & 1436.1641 & August & 65 to 74 years & Male & Widowed\\
\bottomrule
\insertTableNotes
\end{longtable}
\end{ThreePartTable}
\endgroup{}

@@

\begin{ThreePartTable}
\begin{TableNotes}
\item \textit{Note: } 
\item Legend: `PUMFID`: record identification. `EPINO`: episode number. WGHT\_EPI: episode's weight. TUI\_01: activity code. DURATION: episode's duration. LOCATION: episode's location.
\end{TableNotes}
\begin{longtable}[t]{rrlrlll}
\caption{\label{tab:gss-processed-file-2015}\label{tab:ep-2015-processed}Visualization of the active travel episode for the record number `10041` of the 2015 GSS survey.}\\
\toprule
PUMFID & WGHT\_EPI & Activity & Duration & Origin & Destination & Mode\\
\midrule
10041 & 1353.818 & Transport to or from activity & 15 & Home & Home & Walking\\
\bottomrule
\insertTableNotes
\end{longtable}
\end{ThreePartTable}
\endgroup{}

\hypertarget{descriptive-statistics}{%
\subsection{Descriptive statistics}\label{descriptive-statistics}}

Considering GSS Cycles analyzed, we identified 21,748 episodes that
recorded active travel episodes, with trip duration ranging from 0 to
900 minutes, to twelve different destinations. \{ActiveCA\} includes all
these episodes ready for analysis. Table \ref{tab:table-01} presents
descriptive statistics on walking and cycling trips between 1986 and
2015, with measures of trip duration in minutes. The 1986 survey did not
include bicycle trips.

@@

\begin{longtable}[t]{>{}llcccccc}
\caption{\label{tab:table-01}\label{tab:table-01}Descriptive statistics for episodes with active transport records}\\
\toprule
\multicolumn{2}{c}{ } & \multicolumn{6}{c}{Year} \\
\cmidrule(l{3pt}r{3pt}){3-8}
Mode & Statistic & 1986 & 1992 & 1998 & 2005 & 2010 & 2015\\
\midrule
 & Maximum & 660 & 300 & 255 & 515 & 480 & 900\\
\nopagebreak
 & Mean & 21 & 21 & 12 & 12 & 13 & 18\\
\nopagebreak
 & Median & 15 & 10 & 5 & 10 & 10 & 10\\
\nopagebreak
 & Minimum & 1 & 1 & 1 & 0 & 0 & 5\\
\nopagebreak
\multirow[t]{-5}{*}{\raggedright\arraybackslash \textbf{Walking}} & Standard deviation & 31 & 25 & 17 & 16 & 17 & 27\\
\cmidrule{1-8}\pagebreak[0]
 & Maximum &  & 240 & 90 & 180 & 153 & 120\\
\nopagebreak
 & Mean &  & 28 & 24 & 20 & 19 & 25\\
\nopagebreak
 & Median &  & 15 & 15 & 15 & 10 & 20\\
\nopagebreak
 & Minimum &  & 5 & 2 & 1 & 1 & 5\\
\nopagebreak
\multirow[t]{-5}{*}{\raggedright\arraybackslash \textbf{Cycling}} & Standard deviation &  & 36 & 18 & 18 & 23 & 20\\
\bottomrule
\end{longtable}
\endgroup{}

@@

\begin{ThreePartTable}
\begin{TableNotes}
\item \textit{Note: } 
\item 'Min' denotes the minimum time to reach the destination; 'Max' denotes the maximum time to reach the destination; '(\%)' indicates a percentage of the total time to reach the destination; 'Med' refers to the median time to reach the destination
\end{TableNotes}
\begin{longtable}[t]{ccccc>{}c|ccc>{}c|cccc}
\caption{\label{tab:table-02}\label{tab:table-02}Comparison of travel statistics by mode and destination: 1986, 1992, 1998}\\
\toprule
\multicolumn{2}{c}{ } & \multicolumn{4}{c}{1986} & \multicolumn{4}{c}{1992} & \multicolumn{4}{c}{1998} \\
\cmidrule(l{3pt}r{3pt}){3-6} \cmidrule(l{3pt}r{3pt}){7-10} \cmidrule(l{3pt}r{3pt}){11-14}
\multicolumn{1}{c}{\textbf{Destination}} & \multicolumn{1}{c}{\textbf{Mode}} & \multicolumn{1}{c}{\textbf{Min}} & \multicolumn{1}{c}{\textbf{Med}} & \multicolumn{1}{c}{\textbf{Max}} & \multicolumn{1}{c}{\textbf{(\%)}} & \multicolumn{1}{c}{\textbf{Min}} & \multicolumn{1}{c}{\textbf{Med}} & \multicolumn{1}{c}{\textbf{Max}} & \multicolumn{1}{c}{\textbf{(\%)}} & \multicolumn{1}{c}{\textbf{Min}} & \multicolumn{1}{c}{\textbf{Med}} & \multicolumn{1}{c}{\textbf{Max}} & \multicolumn{1}{c}{\textbf{(\%)}}\\
\midrule
 & Home &  &  &  &  & 5 & 20 & 240 & 55.7 & 2 & 15 & 90 & 51.6\\
\nopagebreak
 & Other's home &  &  &  &  & 5 & 10 & 145 & 19.7 & 2 & 10 & 80 & 15.7\\
\nopagebreak
\multirow[t]{-3}{*}{\centering\arraybackslash Cycling} & Work or school &  &  &  &  & 5 & 15 & 45 & 24.7 & 5 & 25 & 75 & 32.8\\
\cmidrule{1-14}\pagebreak[0]
 & Home & 1 & 15 & 330 & 46.3 & 1 & 15 & 300 & 60.5 & 1 & 5 & 255 & 51.9\\
\nopagebreak
 & Other's home & 1 & 10 & 660 & 42.6 & 1 & 5 & 135 & 18.7 & 1 & 5 & 120 & 27.8\\
\nopagebreak
\multirow[t]{-3}{*}{\centering\arraybackslash Walking} & Work or school & 1 & 10 & 450 & 11.1 & 2 & 10 & 60 & 20.9 & 1 & 10 & 75 & 20.3\\
\bottomrule
\insertTableNotes
\end{longtable}
\end{ThreePartTable}
\endgroup{}

@@

\{ActiveCA\} also enables obtaining insights from the main processed
files. Figure \ref{fig:figure-stress} present how the level of stress
varied among respondents depending on their marital status in 2015.
According to this plot, married respondents reported the highest level
of stress, relating to feel stressed every day, with 17\% of possible
cases.
